\begin{abstract}
  离线强化学习通过从预先收集的静态数据集中学习策略，避免了在线交互的高昂成本与安全风险，在医疗诊断、自动驾驶、机器人控制等实际应用场景中具有重要价值。然而，离线数据集的有限覆盖范围导致的分布偏移问题是制约该技术发展的核心挑战：当策略访问离线数据未覆盖的状态-动作对时，价值函数的估计将产生外推误差，进而导致策略性能退化。本文围绕离线强化学习中的不确定性估计与探索策略展开系统研究，提出两种互补的算法方案。

  第一，提出基于自适应不确定性度量的离线强化学习算法。针对现有基于模型方法仅考虑状态转移不确定性、采用静态惩罚系数的局限，引入累积状态不确定性度量和自适应动态权重机制。累积状态不确定性通过递推公式综合考虑历史不确定性的累积效应，使智能体能够更全面地评估决策风险；自适应动态权重机制基于Sigmoid函数动态调节惩罚强度，实现探索与保守的自适应平衡。理论分析证明累积不确定性的单调性和自适应权重的优越性。实验结果表明，该方法在D4RL MuJoCo基准测试集上显著优于MOPO等基线方法。

  第二，提出基于定向探索模型的离线到在线强化学习算法。针对现有离线到在线方法不确定性利用不充分、在线交互成本高的问题，将不确定性信息从"惩罚信号"转变为"探索引导信号"。在离线阶段通过集成模型进行保守策略优化，在在线阶段显式引导策略探索高不确定性且高价值的区域。提出双重不确定性估计方法区分偶然不确定性和认知不确定性，设计动态权重自适应机制平滑过渡探索与利用。理论分析证明定向探索策略相比均匀采样具有更优的收敛速率。实验结果表明，该方法仅需50k步在线交互即可显著优于现有基线方法。

  两种方法形成递进关系：第一种方法解决纯离线设定下的保守策略优化问题，第二种方法通过定向探索策略突破离线数据边界，实现策略性能的持续提升。


  \keywords{离线强化学习，不确定性估计，分布偏移，定向探索，离线到在线强化学习}

\end{abstract}

















\begin{englishabstract}

  Offline reinforcement learning enables policy learning from pre-collected static datasets, avoiding the high costs and safety risks associated with online interactions. This paradigm holds significant practical value in domains such as medical diagnosis, autonomous driving, and robotic control. However, the distribution shift problem arising from the limited coverage of offline datasets remains a core challenge: when the learned policy visits state-action pairs not covered by the offline data, value function estimation produces extrapolation errors, leading to policy performance degradation. This thesis conducts a systematic study on uncertainty estimation and exploration strategies in offline reinforcement learning, proposing two complementary algorithmic approaches.

  First, an offline reinforcement learning algorithm based on adaptive uncertainty measurement is proposed. To address the limitations of existing model-based methods that only consider state transition uncertainty and employ static penalty coefficients, this work introduces cumulative state uncertainty measurement and an adaptive dynamic weight mechanism. The cumulative state uncertainty comprehensively accounts for the accumulated effects of historical uncertainty through a recursive formula, enabling the agent to more thoroughly assess decision risks. The adaptive dynamic weight mechanism based on the Sigmoid function dynamically adjusts penalty strength, achieving an adaptive balance between exploration and conservatism. Theoretical analysis proves the monotonicity of cumulative uncertainty and the advantages of adaptive weights. Experimental results demonstrate that this method significantly outperforms baseline methods such as MOPO on the D4RL MuJoCo benchmark.

  Second, an offline-to-online reinforcement learning algorithm based on directed exploration model is proposed. To address the insufficient utilization of uncertainty and high online interaction costs in existing offline-to-online methods, this work transforms uncertainty information from a ``penalty signal'' to an ``exploration guidance signal''. During the offline phase, conservative policy optimization is performed through ensemble models; during the online phase, the policy is explicitly guided to explore regions with high uncertainty and high action value. A dual uncertainty estimation method is proposed to distinguish aleatoric and epistemic uncertainty, and a dynamic weight adaptation mechanism is designed to smoothly transition between exploration and exploitation. Theoretical analysis proves that the directed exploration strategy achieves better convergence rates compared to uniform sampling. Experimental results show that this method significantly outperforms existing baseline methods with only 50k steps of online interaction.

  The two methods form a progressive relationship: the first addresses conservative policy optimization in pure offline settings, while the second breaks through the boundaries of offline data through directed exploration strategies, achieving continuous improvement in policy performance.


  \englishkeywords{Offline Reinforcement Learning, Uncertainty Estimation, Distribution Shift, Directed Exploration, Offline-to-Online Reinforcement Learning}

\end{englishabstract}
